Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-08-02 paint-contour-tree.tex
small changes

\begin{corollary}\label{cor:geometric}
 Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$.
 Then for any $w \in d$ and $v \in \lambda(p)$, $|H_w| = O(|H_v|)$.
\end{corollary}

\begin{proof} By the previous lemma, $|H_w| = O(|H_p|)$. Note that there are at most
$|H_p|/100$ nodes in $\lambda(p)$, since these are all in the tall support chain of $v$
(which is unsupported). Thus, for any $v \in \lambda(p)$, $H_v \geq 99|H_p|/100$,
completing the proof.
\end{proof}

@@

\begin{definition}\label{def:buff}
 Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
  \begin{enumerate}
  \item $\sum_{p\in \buff} |\loss(p)| > \frac{1}{100} \sum_{p\in \buff} |p|$
  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
 \end{enumerate}
\end{definition}

@@

\begin{proof} Take any reverse topological order of the nodes in $\cU$
(so descendants occur before ancestors). Initialize $\cU'$ to $\cU$.
We process the nodes in order. For node $p$, let $\cU'_p$ denote the rooted
subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(p)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$. (Note that $\cU'$ is redefined
to be the remainder.)

@@

\Lem{cu-root} asserts the root $r$ in $\pruned$ pretty much encompasses all sizes and heaps in $\pruned$.
Observe that part (i) of \Lem{cu-root} is immediate, as it is a special case of \Cor{geometric}. Proving part (ii) requires much more work.
% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
% The reduced heap $\redH_p$ is the
% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
% for a child $q'$ in $\cU$.

@@

\begin{proof} We define $\alpha$ inductively, from leaves upwards.
For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider an internal node $p$
with children $q_1, q_2, \ldots, q_k$. The heap $H(p)$ is formed by the union
of the $H(q_i)$ heaps, after deleting certain vertices from them, and potentially
adding some new vertices. The new vertices are potentially those that come
from other nodes that have been pruned off.
The vertices in each $q_i$ as well as $\lambda(q_i)$ are deleted.
Define $\alpha_p$ to be the number of new vertices added to $H_p$.

Thus, we have $|H_p| = \sum_{i \leq k} |H_q| + \alpha_p$
$- \textrm{\# vertices deleted from $H_q$s}$.
The number of vertices deleted is at most $2\sum_{i \leq k} (|q_i| + |\lambda(q_i)|)$.
There is a factor $2$ because each vertex appears in at most $2$ heaps.

@@

After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\pruned_p$,
denoted $a_0, a_1, a_2, \ldots, a_k$, contributes to the charge at $q$. The charge is expressed in the following
equation. We use the assumption that $A_{a_i} \geq (5/4) A_{a_{i-1}}$ and hence $A_{a_i} \geq (5/4)^i A_{a_0} \geq (5/4)^i$, 
as $a_0$ is an unsupported short path and hence $A_{a_0}\geq 1$. 
$$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (5/4)^{i/2} \leq \alpha_q/20 $$
The total charge is $\sum_{q \in \pruned_p} \alpha_p/20 = A_p/20$. 
\end{proof}

@@

\begin{proof} We prove this by induction over the depth of $q$.
When $q$ is a leaf, this is vacuously tree. Consider some internal node $q$,
with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
hypothesis to the subshrubs rooted at $q'_i$, we can invoke \Clm{charge}
on each subshrub $\pruned_{q'_i}$. Thus, for all $i$,
$\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
By \Clm{weights}, $|H_{q'_i}| \leq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
to complete the proof.
\end{proof}

@@

By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r$.

>>>>

\begin{corollary}\label{cor:geometric}
 Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$, and $p$ 
 is not the root of $\cU$.  Then for any $w \in d$ and $v \in \lambda(p)$, $|H_w| = O(|H_v|)$.
\end{corollary}

\begin{proof} By the previous lemma, $|H_w| = O(|H_p|)$. Note that there are at most
$|H_p|/100$ nodes in $\lambda(p)$, since these are all in the tall support chain of $p$
(which is unsupported). Thus, for any $v \in \lambda(p)$, $|H_v| \geq 99|H_p|/100$,
completing the proof.
\end{proof}

@@

\begin{definition}\label{def:buff}
 Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub}, if it does not contain the root of $\cU'$ (i.e.\ it is a proper subtree), and the following conditions hold:
  \begin{enumerate}
  \item $\sum_{p\in \buff} |\loss(p)| > \frac{1}{100} \sum_{p\in \buff} |p|$
  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
 \end{enumerate}
\end{definition}

@@

\begin{proof} Take any reverse topological order of the nodes in $\cU$
(so descendants occur before ancestors). Initialize $\cU'$ to $\cU$.
We process the nodes in order. For node $p$, let $\cU'_p$ denote the rooted
subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(p)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$, unless $p$ is the root. 
(Note that $\cU'$ is redefined
to be the remainder.)

@@

\Lem{cu-root} asserts the root $r$ in $\pruned$ pretty much encompasses all sizes and heaps in $\pruned$.
Observe that part (i) of \Lem{cu-root} is immediate, as it follows directly from \Lem{geometric}. Proving part (ii) requires much more work.
% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
% The reduced heap $\redH_p$ is the
% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
% for a child $q'$ in $\cU$.

@@

\begin{proof} We define $\alpha$ inductively, from leaves upwards.
For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider an internal node $p$
with children $q_1, q_2, \ldots, q_k$. The heap $H_p$ is formed by the union
of the $H_{q_i}$ heaps, after deleting certain vertices from them, and potentially
adding some new vertices. The new vertices are potentially those that come
from other nodes that have been pruned off.
The vertices in each $q_i$ as well as $\lambda(q_i)$ are deleted.
Define $\alpha_p$ to be the number of new vertices added to $H_p$.

Thus, we have $|H_p| = \sum_{i \leq k} |H_q| + \alpha_p$
$- \textrm{\# vertices deleted from $H_q$'s}$.
The number of vertices deleted is at most $2\sum_{i \leq k} (|q_i| + |\lambda(q_i)|)$.
There is a factor $2$ because each vertex appears in at most $2$ heaps.

@@

After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\pruned_p$,
denoted $a_0, a_1, a_2, \ldots, a_k$, contributes to the charge at $q$. The charge is expressed in the following
equation. We use the assumption that $A_{a_i} \geq (5/4) A_{a_{i-1}}$ and hence $A_{a_i} \geq (5/4)^i A_{a_0} \geq (5/4)^i$, 
as $a_0$ is an unsupported short path and hence $A_{a_0}\geq 1$. 
$$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (4/5)^{i/2} \leq \alpha_q/20 $$
The total charge is $\sum_{q \in \pruned_p} \alpha_p/20 = A_p/20$. 
\end{proof}

@@

\begin{proof} We prove this by induction over the depth of $q$.
When $q$ is a leaf, this is vacuously true. Consider some internal node $q$,
with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
hypothesis to the subshrubs rooted at $q'_i$, we can invoke \Clm{charge}
on each subshrub $\pruned_{q'_i}$. Thus, for all $i$,
$\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
By \Clm{weights}, $|H_{q'_i}| \leq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
to complete the proof.
\end{proof}

@@

By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r|$.
